# Extending Beyond Regression Means

People generally understand the concept of the arithmetic mean. You see it some time during elementary school, it gets tossed around in daily language (usually using the word "average"), and it is statistically important. After all, where would the normal distribution be without a mean? Why, though, do we feel so tied to it from a regression modeling perspective? Yes, it has handy features, but it is also a bit restrictive to the types of relationships that it can actually model well. 

In this chapter, we want to show you what to do when the mean betrays you -- and trust us, the mean will betray you at some point. 

## Quantile Regression

> Oh, you think the mean is your ally. But you merely adopted the mean; I was born in it, molded by it. I didn’t see anything interesting until I was already a man. And by then, it was nothing to me but illuminating.
> -- Bane (probably)
  
```{r, include = FALSE}
# source("load_packages.R")
# source("setup.R")

reticulate::use_condaenv("book-of-models")
```

In a perfect data world, the mean is equal to the middle observation of the data: the *median*. That is only in the perfect world, though, and usually our data comes loaded with challenges. Extreme scores in your data will cause a rift between the median and the mean. 

Let's say we take the integers between 1 and 10, and find the mean. 

$$\frac{1+2+3+4+5+6+7+8+9+10}{10} =  5.5$$

The middle value in that vector of numbers would also be 5.5. 

What happens we replace the 1 with a more extreme value, like -10?

$$\frac{-10+2+3+4+5+6+7+8+9+10}{10} =  4.5$$

With just one dramatic change, our mean went down by a whole point. The median observation, though, is still 5.5. In short, the median is invariant to wild swings out in the tails of your numbers.  

You might be saying to yourself, "Why should I care about this central tendency chicanery?" Let us tell you why you should care -- the least squares approach to the standard linear model dictates that the regression line needs to be fit through the means of the variables. If you have extreme scores that influence the mean, then your regression line will also be influenced by those extreme scores.

Let's look at a few different regression lines:
  
```{r, linear_line_no_extremes}
#| echo: false
#| label: linear_line_no_extremes
#| tbl-cap: Linear line without extreme scores

library(ggplot2)

set.seed(123)
N <- 1000 
k <- 2   
X <- matrix(rnorm(N * k), ncol = k)  
y <- -.5 + .2*X[, 1] + .1*X[, 2] + rnorm(N, sd = .5)  

dfXy <- data.frame(X, y)

ggplot(dfXy, aes(x = X1, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("x") +
  ylab("y") +
  theme_minimal()
```

Now, what would happen if we replaced a few of our observations with extreme scores?
  
```{r linear_line_extremes}
#| echo: false
#| label: linear_line_extremes
#| tbl-cap: Linear line with extreme scores
new_df <- dfXy

new_df$y[dfXy$y > quantile(dfXy$y, .95) & 
           dfXy$X1 > quantile(dfXy$X1, .95)] <- rnorm(10, 2.5, .1)

new_df$X1[dfXy$y > quantile(dfXy$y, .95) & 
            dfXy$X1 > quantile(dfXy$X1, .95)] <- rnorm(10, 2.5, .1)

ggplot(new_df, aes(x = X1, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("x") +
  ylab("y") +
  theme_minimal()
```

With just a casual glance, it doesn't look like our two regression lines are that different. They both look like they have a similar positive slope, so all should be good. To offer a bit more clarity, though, let's put those lines in the same space:
  
```{r, both_lines_plotted}
#| echo: false
#| label: both_linear_lines
#| tbl-cap: Line lines with and without extreme scores

ggplot() +
  geom_smooth(aes(X1, y, color = "No extreme scores"), 
              data = dfXy, method = 'lm', 
              se = FALSE) +
  geom_smooth(aes(X1, y, color = "Extreme scores"), 
              data = new_df, method = 'lm', 
              se = FALSE) +
  xlab("x") +
  ylab("y") +
  theme_minimal() +
  scale_color_manual(name='Data',
                     breaks=c('No extreme scores', 'Extreme scores'),
                     values=c('No extreme scores'='red', 'Extreme scores'='blue'))
```

With 1000 observations, we see that having just 10 extreme scores is enough to change the regression line, even if just a little. There are a few approaches we could take here, with common approaches being dropping those observations or Windsorizing them. Throwing away data because you don't like the way it behaves is nearing on statistical abuse and Windsorization is just replacing those extreme values with numbers that you like a little bit better.

A better answer to this challenge might be to not fit the regression line through the mean, but the median instead. This is where quantile regression becomes handy. Formally, this model can be expressed as:

$$
Q_{Y\vert X}(\tau) = X\beta_\tau
$$

Where we can find the estimation of $\beta_\tau$ as:

$$
\hat\beta_\tau = \arg \min_{\beta \in \mathbb{R}^k} \sum_{i-1}^n(\rho_\tau(Y_i-X_i\beta))
$$

With quantile regression, we are given an extra parameter for the model: $\tau$ or *tau*. The tau parameter let's us choose which quantile we want to use for our line fitting. Since the median splits the data in half, we can translate that to a quantile of .5. 


### Data Import and Preparation

Let's bring in our movie reviews data. Let's say that we are curious about the relationship between the `user_age` variable and the `rating` variable. First, we need to get our data ready for our home-brewed functions.

:::{.panel-tabset}

##### R

```{r, r_data_prep}
#| echo: true
#| label: r_data_prep
reviews <- read.csv("data/review_data.csv")

reviews <- na.omit(reviews)

y <- reviews$rating

X <- cbind(intercept = 1, 
           user_age = reviews$user_age)
```

##### Python

```{python}
import pandas as pd
import numpy as np
reviews = pd.read_csv("~/Documents/book-of-models/data/review_data.csv")
```


```{python}
reviews = reviews.dropna()

y = reviews['rating']

X = pd.DataFrame(
  {'intercept': 1, 
  'user_age': reviews['user_age']}
)
```

### Quantile Loss Function

Let's see one way that we can create a least squares loss function for fitting a linear regression model and compare it with a function for quantile loss.

:::{.panel-tabset}

##### R

```{r, r_lsl_function}
#| echo: true
#| eval: false
#| label: least_squares_loss
least_squares_loss <- function(par, X, y) {
  
  linear_parameters <- X %*% par
  
  mu <- linear_parameters   
  
  loss <- crossprod(y - mu)
}
```

```{r, r_quantile_loss}
#| echo: true
#| label: quantile_loss
quantile_loss <- function(par, X, y, tau) {
  
  linear_parameters <- X %*% par
  
  residual <- y - linear_parameters
  
  loss <- ifelse(residual < 0 , 
                (-1 + tau)*residual, 
                tau*residual)
  
  sum(loss)
}
```

##### Python

```{python}
def least_squares_loss(par, X, y):
# Use matrix multiplication between X and par:
  linear_parameters = X.dot(par)
  
  mu = linear_parameters
  
  loss = np.dot(y - mu, y - mu)
  
  return loss
```

```{python}
def quantile_loss(par, X, y, tau):
# Use matrix multiplication between X and par:
  linear_parameters = X.dot(par)
  
  residual = y - linear_parameters
  
  if residual < 0: 
    loss = (-1 + tau)*residual 
  else: loss = tau*residual
  
  sum(loss)
```


You'll notice right away that we have a few differences. Our quantile loss function includes the *tau* argument, which will let us set our quantile of interest; naturally, it can be any value between 0 and 1. The residual is multiplied by the tau value, only if the residual is greater than 0. If the residual is negative, we need to add tau to -1. Since we need a positive value for our loss values, we will multiply our negative residuals by the negative value produced from -1 plus our tau value. After that, we just sum all of those positive loss values and do our best to minimize that summed value. 

### Model Fitting

Now that we have our data and our loss function, we can fit the model almost exactly like our standard linear model. Again, note the difference here with our tau value, which we've set to .5 to represent the median.

:::{.panel-tabset}

##### R

```{r, r_quantile_optim}
#| echo: true
#| label: r_quantile_optim
optim(
  par = c(intercept = 0, user_age = 0),
  fn  = quantile_loss,
  X   = X,
  y   = y,
  tau = .5
)$par
```

##### Python

```{python}
from scipy.optimize import minimize

minimize(quantile_loss, 
         x0 = np.array([0, 0]), 
         args = (X, y, .5))
```

:::

Fortunately, our interpretation of this result isn't all that different from a standard linear model -- the rating should increase by .08 for every year older that people get. However, this is at the median, not the mean, like the standard linear model. 

### Checking Results

Let's check out results against those produced by `quantreg` and `statsmodels`.

:::{.panel-tabset}

##### R
```{r, r_quantreg}
#| echo: true
#| label: r_quantreg
library(quantreg)

median_test <- rq(rating ~ user_age, tau = .5, 
                data = reviews)

summary(median_test)
```

##### Python

```{python} 
import statsmodels.formula.api as smf

median_test = smf.quantreg('rating ~ user_age', 
                           data = reviews).fit(q = .5)
```

Our home-brewed function returned the same results.

Quantile regression is not a one-trick-pony. Remember, it is called quantile regression -- not median regression. Being able to compute a median regression is just a nice by-product. What we can do with quantile regression is to model different quantiles of the same data. It gives us the ability to answer brand new questions -- does the relationship between user age and their ratings change at different quantiles of rating?

```{r, quantile_lines}
#| echo: false
#| label: quantile_lines
#| fig-cap: Quantile regression lines
tau_values <- c(.1, .3, .5, .7, .9)

quant_values <- purrr::map_df(tau_values, ~{
  result <- coef(rq(rating ~ user_age, tau = .x, 
                    data = reviews))
  result$tau <- .x
  result
})

colnames(quant_values) <- c("intercept", "user_age", "tau")

quant_values$tau <- as.factor(quant_values$tau)

ggplot() +
  geom_point(data = reviews, 
             mapping = aes(user_age, rating), 
             alpha = .5) +
  geom_abline(mapping = aes(slope = user_age, 
                            intercept = intercept, 
                            color = tau),
              data = quant_values) +
  theme_minimal()

```

Instead of a single model to capture the trend through the mean of the data, we can now examine the trends within 5 different quantiles of the data (we aren't limited to just those quantiles, though, and you can examine any of them that you might find interesting). If we had to put some words to our visualization, we could say that all of the quantiles show a positive relationship.  The 30th, 50th, and 70th quantiles all look very close together and appear to have very similar slopes. We can also observe that the 90th quantile appears to be a little flatter than the others, but still positive.

## Additive Models

> Wiggle, wiggle, wiggle, yeah!
> -- LMFAO
  
Fitting a line through your data is always going to be useful, regardless of whether you are using the median or the mean. Those lines give us a wonderful ability to say important things about the relationships between variables and how one variable might influence another. What if we just want to dispense with the notion that we need to fit a straight line through some mass of the data? What if we relax the idea that we need a straight line and think in terms of fitting something curvy through the data.

In other words, we can go from this:
  
```{r, regular_linear_line}
#| echo: false
#| label: regular_linear_line
#| fig-cap: A standard linear model
ggplot(reviews, aes(user_age, rating)) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```

To this:
  
```{r, gam_model_line}
#| echo: false
#| label: gam_model_line
#| fig-cap: A generalized additive model
ggplot(reviews, aes(user_age, rating)) +
  geom_smooth(method = "gam", se = FALSE) +
  theme_minimal()
```

That wiggly line is called a spline. Oddly enough, we can still use a linear model to fit this spline through the data. While this might not give us the same tidy explanation that a typical line would offer, we will certainly get better prediction. 

These models belong to a broad group of *generalized additive models*, often shortened to GAMs. When we fit a quantile regression, we made a slight tweak to the *y*; to fit a GAM, we are going to tweak our predictors. How are we going to tweak them, you might ask? We are essentially going to let them be an *additive* function of features. We will have a model that looks like this:

$$
y = f(x) + \epsilon
$$

These additive features have no concern about linearity with the outcome and will capture nonlinearities in our data very nicely. At this point, you might be asking yourself, "Couldn't I just use some type of polynomial regression or even a nonlinear regression?" Of course you could, but both have pretty serious limitations. The typical polynomial regression likely doesn't fit beyond the data that you currently have and **you** are forcing curves to fit through the data. To use a nonlinear model, you need to know what the underlying nonlinear form actually looks like before you can even specify the model. A GAM will do away with both of these issues; it will produce a curve that will best fit through the data, without the need to know the underlying linear form. 

### Splines

We are going to need to generate several functions to make this work. The first will be to produce the *cubic spline*. Do take note that there are many different types of splines that could be used.

:::{.panel-tabset}

##### R

```{r, r_cubic_spline}
#| echo: true
#| label: r_cubic_spline
cubic_spline <- function(x, z) {
  ((z - 0.5)^2 - 1/12) * ((x - 0.5)^2 - 1/12)/4 -
    ((abs(x - z) - 0.5)^4 - (abs(x - z) - 0.5)^2 / 2 + 7/240) / 24
}
```

##### Python

```{python, python_cubic_spline}
#| echo: true
#| label: python_cubic_spline
import numpy as np

def cubic_spline(x,z):
    return (((z - 0.5)**2 - 1/12) * ((x - 0.5)**2 - 1/12)/4 -
            ((np.abs(x - z) - 0.5)**4 - (np.abs(x - z) - 0.5)**2 / 2 + 7/240) / 24)
```

:::

### Model Matrix Function

Then we a function to produce the model matrix:
:::{.panel-tabset}

##### R
```{r, r_model_matrix}
#| echo: true
#| label: r_model_matrix
splX <- function(x, knots) {
  q <- length(knots) + 2        # number of parameters
  n <- length(x)                # number of observations
  X <- matrix(1, n, q)          # initialized model matrix
  X[ ,2] <- x                   # set second column to x
  X[ ,3:q] <- outer(x, knots, FUN = cubic_spline) 
  X
}
```

##### Python

```{python, python_model_matrix}
#| echo: true
#| label: python_model_matrix

def splX(x, knots):
    q = len(knots) + 2
    n = len(x)
    X = np.ones((n, q))
    X[:,1] = x
    for i in range(2, q):
        X[:,i] = cubic_spline(x, knots[i-2])
    return X
```

This model matrix will help us to produce an *unpenalized spline*. 

:::

### Data Prep

Now that we have our functions, let's get our data ready to go. It is usually a good idea to scale your predictors.

:::{.panel-tabset}

##### R

```{r, r_min_max_scale}
#| echo: true
#| label: r_min_max_scale
x <- reviews$user_age - min(reviews$user_age)
x <- x / max(x)
rating <- reviews$rating
```

##### Python

```{python, python_min_max_scale} 
#| echo: true
#| label: python_min_max_scale
x = reviews['user_age'] - reviews['user_age'].min()
x = x / x.max()
rating = reviews['rating']
```

:::

### Model Matrix

We can create a model with 4 knots -- you can think of knots as places where individual regression lines will get joined together. You can always experiment with more or less knots. Once we have our knots ready, we can create the model matrix.

As soon as you create your `X` object, you should take a look at it. It will be a matrix with 4 columns. The first column will be all 1s, the second column will be the scaled `user_age`, and the last two columns will be the cubic splines.

:::{.panel-tabset}

##### R

```{r, r_knots}
#| echo: true
#| label: r_knots
knots <- 1:4/5
```

```{r, r_model_matrix_spline}
#| echo: true
#| label: r_model_matrix_spline
X <- splX(x, knots)            
```

```{r, model_matrix_head}
#| echo: true
#| label: model_matrix_head
head(X)
```

##### Python

```{python, python_knots}
#| echo: true
#| label: python_knots

knots = np.arange(1, 5) / 5
```

```{python, python_model_matrix_spline}
#| echo: true
#| label: python_model_matrix_spline

X = splX(x, knots)
```

```{python, model_matrix_head}
#| echo: true
#| label: model_matrix_head

X[:5,:]
```

:::

### Model Fitting

Now that we have a model matrix, `X`, we can fit the model. All of the hardwork was done in creating the model matrix and we can just use `lm` or `OLS` to fit the model.
  
:::{.panel-tabset}

##### R
```{r, r_cubic_model_fitting}
#| echo: true
#| label: r_cubic_model_fitting
fit_lm <- lm(rating ~ X - 1)

fit_lm
```

##### Python

```{python, python_cubic_model_fitting}
#| echo: true
#| label: python_cubic_model_fitting
fit_lm = sm.OLS(rating, X).fit()

fit_lm.summary()
```

:::

### Prediction

We can set some prediction values for this model:

:::{.panel-tabset}

##### R
  
```{r, r_predictions}
#| echo: true
#| label: r_predictions
xp <- seq(0, 1, by = .01)
Xp <- splX(xp, knots)  
```

##### Python

```{python, python_predictions} 
#| echo: true

xp = np.arange(0, 1, 0.01)
Xp = splX(xp, knots)
```
:::

While creating those predictions is nice, using them to visualize the model is far more helpful.
  
```{r, spline_viz}
#| echo: false
#| label: spline_viz
#| fig-cap: Visualizing cubic regression spline
ggplot(aes(x = x, y = rating), data = data.frame(x, rating)) +
  geom_point(color = "#FF5500") +
  geom_line(aes(x = xp, y = Xp %*% coef(fit_lm)),
            data = data.frame(xp, Xp),
            color = "#00AAFF") +
  theme_minimal()
```


### Checking Results

It is always worth checking our solution against known functions. Let's see how our model compares to the `mgcv` package in R and the `statsmodels` package in Python. 

:::{.panel-tabset}

##### R
```{r, mgcv_check}
#| echo: true
#| label: mgcv_check
library(mgcv)
summary(gam(rating ~ X-1))
```

##### Python

```{python, python_mgcv_check}  
#| echo: true
#| label: python_mgcv_check
import statsmodels.api as sm

Xp = sm.add_constant(Xp)
fit_gam = sm.GLM(rating, X, family=sm.families.Gaussian()).fit()

fit_gam.summary()
```
:::


Typically, you don't need to go through the whole mess of creating a model matrix by hand and can use additional functions to fit a spline to your regression. In R, you can use the `gam` function from the `mgcv` package. In Python, you can use the `GAM` function from the `pygam` package.

:::{.panel-tabset}

##### R
```{r, r_gam_demo}
#| echo: true
#| label: r_gam_demo
gam_model <- gam(rating ~ s(x, bs = "cr"))

summary(gam_model)
```

##### Python

```{python, python_gam_demo}  
#| echo: true
#| label: python_gam_demo

from pygam import LinearGAM, s

gam_model = LinearGAM(s(0)).fit(x, rating)

gam_model.summary()
```

:::

### Penalized Cubic Spline

Recall that this is an unpenalized cubic spline. If we want to have a finer degree of control over that wiggly line, we can include a *lambda penalty*. 

We'll need to change up our spline function just a bit.

:::{.panel-tabset}

##### R
  
```{r, r_spline_penalty}
#| echo: true
#| label: r_spline_penalty
splS <- function(knots) {
  q <- length(knots) + 2
  S <- matrix(0, q, q) 
  S[3:q, 3:q] <- outer(knots, knots, FUN = cubic_spline)
  S
}
```

##### Python

```{python, python_spline_penalty}
#| echo: true
#| label: python_spline_penalty

def splS(knots):
    q = len(knots) + 2
    S = np.zeros((q, q))
    S[2:, 2:] = np.outer(knots, knots).apply(cubic_spline)
    return S
```

:::

We also need to be able to take the square root of our entire matrix. This is a bit more complicated than it sounds. We need to take the eigenvalue decomposition of the matrix, take the square root of the eigenvalues, and then recombine the matrix.

:::{.panel-tabset}

##### R

```{r, r_matrix_square}
#| echo: true
#| label: r_matrix_square
mat_sqrt <- function(S) {
  d <- eigen(S, symmetric = TRUE)
  rS <- d$vectors %*% diag(d$values^.5) %*% t(d$vectors)
  rS
}
```

##### Python

```{python, python_matrix_square}
#| echo: true
#| label: python_matrix_square

def mat_sqrt(S):
    d = np.linalg.eig(S)
    rS = d.vectors @ np.diag(d.values**.5) @ d.vectors.T
    return rS
```

:::

### Penalized Model Fitting Function

With those functions in hand, we can create the function to fit the entire model.

:::{.panel-tabset}

##### R
  
```{r, r_penalized_fit}
#| echo: true
#| label: r_penalized_fit
prs_fit <- function(y, x, knots, lambda) {
  q  = length(knots) + 2    # dimension of basis
  n  = length(x)            # number of observations
  Xa = rbind(splX(x, knots), mat_sqrt(splS(knots))*sqrt(lambda)) # augmented model matrix
  y[(n + 1):(n+q)] = 0      # augment the data vector
  
  lm(y ~ Xa - 1) # fit and return penalized regression spline
}
```

##### Python

```{python, python_penalized_fit} 
#| echo: true
#| label: python_penalized_fit

def prs_fit(y, x, knots, lamba):
    q = len(knots) + 2
    n = len(x)
    Xa = np.vstack((splX(x, knots), mat_sqrt(splS(knots))*np.sqrt(lamba)))
    y[(n + 1):(n+q)] = 0
    return sm.OLS(y, Xa).fit()
```

:::

Notice again that magic happens in the model matrix, but that we are still just using `lm` or `OLS` to fit the model.

### Penalized Model Fitting

Let's stick with 4 knots and see what happens when we set our lambda to .1:

:::{.panel-tabset}

##### R
  
```{r, r_lambda_1}
#| echo: true
#| label: r_lambda_1
knots <- 1:4/5

fit_penalized <- prs_fit(
  y <- rating,
  x <- x,
  knots <- knots,
  lambda <- .1
) 

Xp <- splX(xp, knots) 
```

##### Python

```{python, python_lambda_1}
#| echo: true
#| label: python_lambda_1

knots = np.arange(1, 5)/5

fit_penalized = prs_fit(
    y = rating,
    x = x,
    knots = knots,
    lamba = .1
)

Xp = splX(xp, knots)
```

There is definitely some wiggly to that line, but it is not as extreme as what we saw with our unpenalized cubic spline. 

:::
```{r, r_lambda_1_viz}
#| echo: false
#| label: r_lambda_1_viz
#| fig-cap: GAM model with lambda set to .1
ggplot(aes(x = x, y = rating), data = data.frame(x, rating)) +
  geom_point(color = "#FF5500") +
  geom_line(aes(x = xp, y = Xp %*% coef(fit_penalized)),
            data = data.frame(xp, Xp),
            color = "#00AAFF") +
  theme_minimal()
```

We can test out what happens at different lambda values:
  
```{r, lambda_value_viz}
#| echo: false
#| label: lambda_value_viz
#| fig-cap: GAM model with different lambda values
plot_data <- purrr::map_df(c(.9, .5, .1, .01, .001), ~{
  fit_penalized = prs_fit(
    y = rating,
    x = x,
    knots = knots,
    lambda = .x
  ) 
  Xp = splX(xp, knots)
  
  results <- data.frame(
    x = xp,
    y = Xp %*% coef(fit_penalized),
    lambda = as.factor(.x)
  )
  
  return(results)
})

ggplot(aes(x = x, y = rating), data = data.frame(x, rating)) +
  geom_point() +
  geom_line(aes(x = x, y = y, color = lambda),
            data = plot_data) +
  theme_minimal()
```

What can we take from this? As lambda values get closer to 1, we see lines that look very similar to a standard linear model. If you recall the our function to fit the model, we multiplied the square root of the matrix by the square root of the lambda value; since the square root of 1 is 1, we wouldn't see anything too interesting happen. As our lambda value gets lower, we see an increasing amount of wiggle happen. 

Naturally, this is a great time to think about how these models would work on new data. As lambda gets smaller, we are fitting our in-sample data much better. How do you think this would fare with unseen data? If you'd say that we would do well with training and horrible on testing, we'd likely agree with you. 

## Mixed Models

As much fun as modeling is, knowing your data is far more important. You can throw any model you want at your data, from simple to fancy, but you can count on disappointment if you don't fundamentally know the structures that live within your data. Let's take a quick look at the following visualizations:

```{r}
#| echo: false
#| label: year-release-rating
#| fig-cap: Linear relationship between year of movie release and rating.
ggplot(reviews, aes(user_age, rating)) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE) +
  theme_minimal()
```

In this visualization, we don't see much of a relationship between user age and ratings. We could probably just stop there, but we might be ignoring something substantial within our data: year of release We might want to ask a question, "Does this relationship work the same way across the different years?"

```{r}
#| echo: false
#| label: year-release-genre-rating
#| fig-cap: Linear relationship between year of movie release and rating, with genre.
ggplot(reviews, aes(user_age, rating, color = as.factor(year_of_release))) +
  # geom_point() +
  geom_smooth(method="lm", se = FALSE) +
  theme_minimal()
```

A very quick examination of the visualization might suggest that the relationship between user age and rating varies significantly over the different release years. Some years show a positive relationship, while some show a negative relationship, and all with various levels of strength.

Just for fun, try to group by release year and summarize it by the mean rating. You'll find that the averages are very different across the years! Then, subtract the overal mean rating from those values. Keep them handy!

Clearly, release year is offering some type of additional information to the model, but how can we incorporate that information into our model? An interaction might come to find at first, but that becomes a tricky interpretation endeavour. Instead, the **mixed model** can be used to incorporate that information into our model, without much hassle and with a huge boost in explanability. 

The term **mixed model** is as vanilla as we can possibly make it,but you might have heard of different flavors of them before. If you've come from the Social Sciences, you might have heard of *Hierarchical Linear Models*. You might have even heard the words *multilevel models* or *mixed-effects models* tossed around before. Maybe you've even be exposed to ideas like *random effects* or *random slopes*. No matter what word you say, they are all instances of a **mixed model**. 

What makes a model a mixed model? The mixed model is characterized by the idea that a model can have *fixed* and *random* effects. Fortunately, you've already encountered *fixed* effects -- those are the features that we have been using in all of our models so far! The *random effect*, though, is typically some type of grouping-based variable that contributes to unique variance in the outcome and we are going to let them vary from the rest of the model. Grouping variables can be actual groups, with the classic example being students in a classroom. Those groups can also be next within larger groups -- students nested within classrooms, nested within schools, nested within districts. These groups can also capture the notion of a repeated measure for an individual, like repeated test scores. 

While we aren't rejecting the idea of a mean with these mixed-models, we are implying that each group (whether it is a group, nested group, or repeated observations from a person) has its own unique mean that can be useful for modeling the target. Formally, we might specify something like this:

$$
rating = b_{\text{0_genre}} + b_{year}*\text{year}
$$

We are explicitly saying that genre has its own unique slope for this model, where $b_{\text{0_genre}} ~ \eta(b_{intercept}, \tau)$, meaning that the random intercepts will be normally distributed and the overall intercept is just the mean of those random intercepts. 

Before we go to modeling our reviews, let's consider an exmaple training program to increase vertical jump, with average vertical increases of 2 inches. That really doesn't sound all that impressive; however, that increase came across 5 distinct groups: NBA players, NFL players, NHL players, MLB players, and data analysts. We can be pretty certain that each group has a very different vertical jump distribution coming into this training program. Given the amount of jumping that NBA players do, this program is unlikely to produce dramatic increases in vertical jump for them. We would probably expect modest gains in the NFL and even greater gains within NHL and MLB players. Where we are going to see the best gains, though, is from data analysts -- they might want to jump to conclusions, but don't need to jump over their computers very often. Now that we know the additional information, can we just look at the average increase and be satisfied? Probably not. Instead, we need to look at the group that we might be in and judge accordingly. A mixed-model is going to let us to model all of this without too much work.

### Model Matrix

Before we can fit our mixed model, we need to create a model matrix. We are going to use the `model.matrix()` function to create our model matrix. We are going to use the `user_age` variable as our fixed effect and the `year_of_release` as our random effect. We are going to use the `factor()` function to make sure that `year_of_release` is treated as a categorical variable. We are also going to use the `-1` to remove the intercept from the model matrix. 

:::{.panel-tabset}

##### R

```{r}
#| echo: true
#| label: mixed_model_prep

X <- model.matrix(~user_age, reviews)
Z <- model.matrix(~factor(reviews$year_of_release) - 1)

colnames(Z) <- paste0("released_", sort(unique(reviews$year_of_release)))

y <- reviews$rating
```

##### Python

```{python}
#| echo: true
#| label: mixed_model_prep

X = reviews[['user_age']]
X = sm.add_constant(X)

Z = pd.get_dummies(reviews['year_of_release'], drop_first=True)

y = reviews['rating']
```

:::

### Likelihood Function

:::{.panel-tabset}

##### R

```{r, r_data_read}
#| echo: true
#| label: mixed_model_ll

mixed_log_likelihood <- function(y, X, Z, theta) {
  tau   = exp(theta[1])
  sigma = exp(theta[2])
  n = length(y)
  
  # evaluate covariance matrix for y
  e  = tcrossprod(Z)*tau^2 + diag(n)*sigma^2
  b  = coef(lm.fit(X, y))
  mu = X %*% b

  ll = mvtnorm::dmvnorm(y, mu, e, log = TRUE)
  -ll
}
```

##### Python

```{python}
#| echo: true
#| label: mixed_model_ll

def mixed_log_likelihood(y, X, Z, theta):
    tau = np.exp(theta[0])
    sigma = np.exp(theta[1])
    n = len(y)
    
    # evaluate covariance matrix for y
    e = np.matmul(Z, Z.T) * tau**2 + np.eye(n) * sigma**2
    b = np.linalg.lstsq(X, y, rcond=None)[0]
    mu = np.matmul(X, b)
    
    ll = stats.multivariate_normal.logpdf(y, mu, e)
    return -ll
```

### Model Fitting

:::{.panel-tabset}

##### R

```{r}
param_init <- c(0, 0)

names(param_init) <- c('tau', 'sigma')

fit <- optim(
  fn  = mixed_log_likelihood,
  X   = X,
  y   = y,
  Z   = Z,
  par = param_init,
  control = list(reltol = 1e-10)
)

fit
```

##### Python

```{python}
#| echo: true
#| label: mixed_model_fit

param_init = np.array([0, 0])

fit = opt.minimize(
    fun=mixed_log_likelihood,
    x0=param_init,
    args=(X, Z, y),
    method='BFGS',
    tol=1e-10
)

fit
```

### Model Checking

Let's see if we can get a match from `lme4` or `statsmodels`.

:::{.panel-tabset}

##### R
```{r}
library(lme4)

fit_mer = lmer(rating ~ user_age + (1 | year_of_release), 
               reviews, 
               REML = FALSE)

summary(fit_mer)
```

##### Python

```{python}
import statsmodels.api as sm

fit_mer = sm.MixedLM.from_formula("rating ~ user_age", reviews, 
                                  groups=reviews["year_of_release"])

fit_mer = fit_mer.fit()

fit_mer.summary()
```

:::

Now that we know our homemade methods can match the others, what does this model really offer us?

```{r}
#| echo: false
library(sjPlot)

sjPlot::plot_model(fit_mer, type = "re") +
  theme_minimal()
```

Remember your group-by and summarize task earlier? Each point is the difference between the year's average and the overall average -- values in blue are higher than the average and values in red are lower than the average. With that, the average rating for 2001 is much better than the global average rating, while 2019 is much worse than the global average rating. 

## Performance Comparisons

Just for giggles, we should see how all of our models perform:

```{r, model_performace_comp}
#| echo: false
#| label: model_performance_comp
#| fig-cap: Comparing model performance with RMSE
library(mgcv)
model_data <- data.frame(rating = rating, 
                         age = x, 
                         year_of_release = year_of_release)
lm_test <- lm(rating ~ age, 
              data = model_data) 
median_test <- rq(rating ~ age, 
                  data = model_data) 
gam_test <- gam(rating ~ 
                  s(age, bs = "cr", fx = FALSE, m = .001), 
                data = model_data) 

fit_mer = lmer(rating ~ user_age + (1 | year_of_release), 
               reviews, 
               REML = FALSE)

knitr::kable(
  data.frame(model = c("standard", "median", "gam", "mixed"), 
             rmse = c(modelr::rmse(lm_test, model_data),
                      modelr::rmse(median_test, model_data),
                      modelr::rmse(gam_test, model_data), 
                      modelr::rmse(fit_mer, model_data))
  )
)
```

Unsurprisingly, gam was the best performer of the bunch. The results of our median regression and standard linear model are pretty similar, with a slight edge to the standard model. This finding may or may not surprise you -- as you spend more time with models, you often encounter situations where the linear model performs just as well as a model that should, in theory, do much better. The linear model is the **the linear model** for a reason!


## Concluding Ramblings

The standard linear model is useful across many different data situations. It does, unfortunately, have some issues when data becomes a little bit more "real". When you have extreme scores or relationships that a standard model might miss, you don't need to abandon your linear model in favor of something more exotic. Instead, you might just need to think about how you are actually fitting the line through your data. 