# The Foundation

> It is the chief characteristic of data science that it works.
― Isaac Asimov (paraphrased)

```{r setup-lm, include = FALSE}
source("load_packages.R")
source("setup.R")
source("functions/utils.R")

reticulate::use_condaenv("book-of-models")
```

Packages needed or useful for this chapter include:

```{r, echo = TRUE, eval = FALSE}
library(tidyverse)
```




## Introducing the Greatest Of All Time

Now that you have some idea of what you're getting into, it's time to dive in! We'll start things off by covering the building block of all modeling, and a solid understanding here will provide you the basis for just about anything that comes after, no matter how complex it gets. The **linear model** is our starting point. At first glance, it may seem like a very simple model, but it's actually quite powerful and flexible, able to take in different types of inputs, handle nonlinear relationships, temporal and spatial relations, clustering, and more. Linear models have a long history, with even the formal and scientific idea behind correlation and linear regression being well over a century old[^corbypeirce]! And in that time, the linear model is far and away the most used model out there. But before we start talking about the *linear* model, we need to talk about what a **model** is in general.


[^corbypeirce]: Peirce & Bowditch were well ahead of Pearson and Galton [@rovine2004peirce].


### What is a Model?

At its core, a model is just an **idea**. It's a way of thinking about the world, about how things work, how things change over time, how things are different from each other, and how they are similar. The underlying thread here is that **a model expresses relationships** about things in the world around us. One can also think of a **model as a tool**, one that allows us to take information, in the form of data, and act on it in some way. Just like other ideas (and tools), models have consequences in the real world, and they can be used wisely or foolishly.  

On a practical level, a model is expressed through a particular language, math, but don't let that worry you if you're not so inclined. As it's still just an idea at its core, the idea is the most important thing to understand about a model. The **math is just a way of expressing the idea** in a way that can be communicated and understood by others in a standard way, as well as helping make the idea precise. But in everyday terms, we're trying to understand things like how the amount of sleep relates to cognitive functioning, how the weather affects the number of people who visit a park, how much money to spend on advertising to increase sales, how to detect fraud, and so on.  Any of these  could form the basis of a model, as they stem from scientifically testable ideas, and they all express relationships between things we are interested in, possibly even with an implication of causal relations. 

If you wanted to run a linear model to understand the relationship between sleep and cognitive functioning, you might express it in code as:

:::{.panel-tabset}

##### R

```{r}
#| eval: false
lm(cognitive_functioning ~ sleep)
```

##### Python

```{python}
#| eval: false
from statsmodels.formula.api import ols

model = ols('cognitive_functioning ~ sleep', data=df).fit()
```

:::


Very easy! But that's all it takes to express a simple linear relationship.  In this case, we're saying that cognitive functioning is a linear (function) of sleep.  By the end of this chapter you'll also know why R's function is `lm` (linear model) and the [statsmodels]{.pack} function is `ols`, but both are doing the same thing. 

<!-- 
We can also express this as a simple equation:

$$ 
\textrm{cog\_func} = \beta_0 + \beta_1 \cdot \textrm{sleep}
$$

But we'll get more into that later. -->


## Key ideas

We can pose a few concepts key to understanding models. This is not an exhaustive list, but it's a good start.  We'll cover each of these in turn.

- What a model is: The model as an idea
- Features, targets, and input-output mappings: how do we get from input to output?
- Model estimation: how do we find the best model?
- Prediction: how do we use a model?
- Assumptions, probabilistic outcomes, uncertainty: how do we know if we can trust a model?

As we go along and cover these concepts, be sure that you feel you have the 'gist' of what we're talking about.  Almost everything of what comes after linear models builds on these ideas, so it's important to have a firm grasp before climbing to new heights.


## What goes into a model? Features and Targets

In the context of a model, how we specify the nature of the relationship depends on the context. In the interest of generality, we'll refer to the **target** as what we want to explain, and **features** as those aspects of the data we will use to explain it. Because people come at data from a variety of contexts, they often use different terminology to mean the same thing. Some of these actually suggest a particular type of relationship (e.g., a causal relationship, an experimental setting), but we'll ignore that for now.  The table below shows some of the common terms used to refer to features and targets.

```{r tbl-feat-target, cache=FALSE}
#| echo: false
#| label: tbl-feature-target-names
#| tbl-cap: Common Terms for Features and Targets


tbl_feat_targ = tibble(
    Feature = c("independent variable", "predictor variable", "explanatory variable", "covariate", "x", "input", "right-hand side"),
    Target  = c("dependent variable", "response", "outcome", "label", "y", "output", "left-hand side"),
) |>
    gt() |>
    rm_caption() # does nothing

tbl_feat_targ
```

We may use any or all of these words to describe things so that you are comfortable with the terminology, but we'll stick with **features** and **targets** for the most part.  In our opinion, this terminology has the least hidden assumptions/implications.


### Expressing Relationships


As noted,  a model is a way of expressing a relationship between a set of features and a target, and one way of expressing them is as inputs and outputs. But how can we go from input to output?  Well to begin, we assume that the features and target are **correlated**, i.e. that there is some relationship between the x and y.  If so, then we can ulitmately use the features to **predict** the target.  In the simplest setting a correlation implies a  relationship where x and y typically move up and down together (left plot) or they move in opposite directions where x goes up and y goes down.

```{r corr-plot, cache.rebuild=TRUE}
#| echo: false
#| label: fig-corr-plot
#| fig-cap: Correlation

p_dat = tibble(
    x = rnorm(50),
    y = .75 * x + rnorm(50, sd = .5),
    yneg = -.75 * x + rnorm(50, sd = .5)
)

p1 = ggplot(p_dat, aes(x, y)) +
    geom_point() +
    labs(subtitle = "Positive Correlation")

p2 = ggplot(p_dat, aes(x, yneg)) +
    geom_point() +
    labs(subtitle = "Negative Correlation")

p1 + p2
```

In addition, the typical correlation suggests a linear relationship.  There are many types of correlation metrics, but the most common one, the Pearson correlation, is explicitly a measure of the linear relationship between two variables.  It's expressed as a number between -1 and 1, where 0 means there is no linear relationship. As we move closer to 1, we would see a tighter scatterplot like the one on the left, until it became a straight line.  The same happens for the negative relationship as we get closer to a value of -1.  If we have only one feature and target, the Pearson correlation reflects the exact result of the linear model we'd conduct.  Even with multiple features, we often lean a version of the Pearson R to help us understand how the features account for the target's variability.

```{r fig-corr-line-plot}
#| echo: false
p1 = p1 + geom_smooth(method = "lm", se = FALSE)
p2 = p2 + geom_smooth(method = "lm", se = FALSE)
p1 + p2
```



## *THE* Linear Model


The linear model is perhaps the simplest *functional* model we can use to express a relationship between features and targets.  It's possibly still the most common model used in practice, and it is the basis for many types of models.  Why don't we run one now?

The following dataset contains movie individual [movie reviews][app-data-review] and contains the rating (1-5 stars scale) along with features pertaining to the review (e.g. gender, education) and features about the movie (e.g. genre, year).  We'll use the linear model to predict the rating from the length of the review.

```{r}
#| echo: false
#| eval: false
#| label: import-review
library(tidyverse)

df_reviews <- read_csv("data/review_data.csv")

skimr::skim(df_reviews)
```

For our first linear model, we'll keep things simple.  Let's predict the rating from the length of the review.  We'll use the `lm()` function in R and the `ols()` function in Python to fit the model.  The `lm()` function takes a formula as its first argument, which is a way of expressing the relationship between the features and target.  The formula is expressed as `y ~ x`, where `y` is the target and `x` is the feature.  The `ols()` function takes the same formula, but it also requires a `data` argument that specifies the data frame containing the features and target.

:::{.panel-tabset}
##### R

```{r my-first-model}
#| label: r-my-first-model
df_reviews <- read_csv("data/review_data.csv")

model_reviews = lm(rating ~ review_length, data = df_reviews)

summary(model_reviews)
```

```{r}
#| label: r-my-first-model-output

# gt(broom::tidy(model_reviews))
```



##### Python

```{python}
#| label: py-my-first-model
import pandas as pd
import numpy as np
import statsmodels.formula.api as smf

df_reviews = pd.read_csv('data/review_data.csv')
model_reviews = smf.ols('rating ~ review_length', data = df_reviews).fit()

print(model_reviews.summary())
```

:::

For such a simple model, we certainly get a lot of output! Don't worry, you'll eventually come to know what it all means. But it's nice to know how easy it is to get the results!

The linear model posits a **linear combination** of the features.  A linear combination is just a sum of the features, each of which has been multiplied by some specific value.  That value is often called a **coefficient** or possibly **weight**.  The linear model is expressed as (math incoming!):

$$y = b_0 + b_1x_1 + b_2x_2 + ... + b_nx_n$$

- $y$ is the target.
- $x_1, x_2, ... x_n$ are the features.
- $b_0$ is the intercept, which is kind of like a baseline value or offset. If we has no inputs at all it would just be the mean of the target.
- $b_1, b_2, ... b_n$ are the coefficients or weights for each feature.

But lets start with something simpler, let's say you want to take a sum of several features. In math you would write it as:

$$x_1 + x_2 + ... + x_n$$

In the previous equation, x is the feature and n is the number of features. $x_1$ is the first feature, $x_2$ the second, and so on.  $x$ is an arbitrary designation, you could use any letter, symbol you want, or better would be the actual name. Now look at the linear model.


$$y = x_1 + x_2 + ... + x_n$$


In this case, the function is *just a sum*, something so simple we do it all the time. In the linear model sense though, we're actually saying a bit more. Another way to understand that equation is that *y is a function of x*.  We don't show any coefficients, but technically it's as if each coefficient was a value of 1.  In other words, for this simple linear model, we're saying that each feature contributes equally to the target. 

In practice, features will not contribute in the same ways. If we want to relate some feature, x1, and some other feature, x2, to target y, we probably would not assume that they both contribute in the same way from the beginning.  We might give more weight to x1 than x2.  In the linear model, we express this by multiplying each feature by a different coefficient. So the linear model is really just a sum of the features multiplied by their coefficients. In fact, we're saying that each feature contributes to the target in proportion to the coefficient. So if we have a feature x1 and a coefficient b1, then the contribution of x1 to the target is b1\*x1. If we have a feature x2 and a coefficient b2, then the contribution of x2 to the target is b2\*x2. And so on.  So the linear model is really just a sum of the features multiplied by their coefficients.  

For our model, here is the mathemtical represenation:

$$\textrm{rating} = b_0 + b_1 \cdot \textrm{review\_length}$$

And with the actual results of our model:

$$\textrm{rating} = -1.2 + .09 \cdot \textrm{review\_length}$$

Not too complicated we hope!  But let's make sure we see what's going on here just a little bit more. 

- Our *idea* is that the length of the review is in some way related to the eventual rating given to the movie. 
- Our *target* is rating, and the *feature* is the review length
- We *map the feature to the target* via the linear model, which provies an initial understanding of how the feature is related to the target.  In this case, we start with a baseline of -1.2.  This makes sense only in the case of a rating with no review, which is required for this data.  We'll talk about ways to get a more meaningful intercept later, but for now, that is our starting point.
 if we add a single character to the review, we expect the rating to increase by .09. If we add 10 characters, we expect the rating to increase by .9. If we add 20 characters, we expect the rating to increase by 1.8.


> In chapter 0, maybe say something about science = prediction, but maybe also stay away from academic treament

::: {.callout-note collapse="true" appearance="minimal"}
## Matrix Represenation of a Linear Model

Here we'll show the matrix represenation form of the linear model, for the typical case where we have more than one feature in the model. In the following, y is a vector of all target observations, and likewise each x is a (row) vector of all observations for that feature.  The b vector is the vector of coefficients.  The 1 serves as a means to incorporate the intercept. It's just a feature that always has a value of 1.  The matrix multiplication is just a compact way of expressing the sum of the features multiplied by their coefficients.  We can even do it more

Here is y as a vector of observerations, n x 1.

$$
\textbf{y} = \begin{bmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{bmatrix}
$$ {#eq-lm-mat-y}

Here is the vector for x, including the intercept:

$$
\textbf{X} = \begin{bmatrix}
1 & x_{11} & x_{12} & \dots & x_{1p} \\
1 & x_{21} & x_{22} & \dots & x_{2p} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & x_{n1} & x_{n2} & \dots & x_{np}
\end{bmatrix}
$$ {#eq-lm-mat-x}

And finally, here is the vector of coefficients:

$$
\textbf{b} = \begin{bmatrix}
b_0 \\
b_1 \\
\vdots \\
b_p
\end{bmatrix}
$$ {#eq-lm-mat-b}


Putting it all together, we get the linear model in matrix form:

$$
\textbf{y = Xb }
$$ {#eq-lm-mat-mult}

:::

### Intepretation

```{r}
#| echo: false
#| label: my-first-model-coefs-etc


intercept = round(coef(model_reviews)[1], 2)
rl_coef = round(coef(model_reviews)[2], 2)
sd_y = round(sd(df_reviews$rating), 2)
sd_x = round(sd(df_reviews$review_length), 2)
rl_ci = round(confint(model_reviews)[2, ], 2)
n_char = 50
```

Let's try and use some context to interpret the coefficient. The size for the coefficient is `r round(rl_coef, 2)`, and the standard deviation of the target, i.e. how much it moves around naturally on its own, is `r sd_y`.  So the coefficient is about `r round(rl_coef / sd_y * 100, 0)`% of the standard deviation of the target.  In other words, a single character addition to a review results in an expected increase of `r round(rl_coef / sd_y * 100, 0)`% of what the review would normally bounce around in value. Most would think this is not negligible, and by some standards maybe even large [CITATION NEEDED- COHEN?].  Why? Because that's just for a single character increase in the review length.  Well, what would be a notable increase in review length?  An easy and straightforward measure would be the standard deviation of the feature. In this case it's `r sd_x`.  So if we increase the review length by one standard deviation, we expect the rating to increase by `r round(rl_coef * sd_x, 2)`, which is `r  round(sd_x/sd_y * rl_coef, 2)` of the standard deviation of the target.  That's a pretty big jump!  So we can see that the coefficient is probably not what we'd call negligible, and that the feature is indeed related to the target.  But we can also see that the coefficient is not so large that it's not believable.

::: {.callout-tip}
The calculation we just did results in what's often called a 'standardized' coefficient.  In the case of the simplest model with one feature like this one, it is identical to the Pearson r correlation metric, which we invite you to check and confirm on your own. In the case of multiple features, it is the correlation between the target and the feature, after adjusting for the other features.  But before you start thinking of it as a measure of *importance*, it is not. It provides some measure of the feature-target linear relationship, but that doesn't not entail practical importance, nor assess nonlinear relationships, interactions, and a host of other interesting things.
:::


## What do we do with a model? Prediction

Once we have a working model, we can use it to make predictions.  We can do this by plugging in values for the features and using the corresponding weights. Let's go back to our results, starting with a simpler depiction.

```{r}
#| echo: false
#| label: my-first-model-output

broom::tidy(model_reviews, conf.int = TRUE) |>
    janitor::clean_names() |>
    rename(feature = term) |>
    mutate(feature = ifelse(feature == "(Intercept)", "intercept", feature)) |>
    gt()
```

The table shows the **coefficient** for each feature including the intercept.  In this case, the coefficient is `r rl_coef`, which means that for every additional word in the review, the rating goes up by `r rl_coef` stars.  So if we had a review that was `r n_char` characters long, we would predict a rating of `r paste(intercept)` + `r glue("{n_char}*{rl_coef}")` = `r round(predict(model_reviews, newdata = data.frame(review_length = n_char)), 1)` stars.

When we're talking about predictions we usually will see this as:

$$\hat{y} = b_0 + b_1x_1 + b_2x_2 + ... + b_nx_n$$

What is $\hat{y}$? The hat over the $y$ just means that it's a predicted value of the model, rather than the one we actually observe. In fact, we were missing something in our previous depictions of the linear model.  We need to add an error term, $\epsilon$, to account for the fact that our predictions will not be perfect[^perfect_prediction].  So the full linear model is:

[^perfect_prediction]: In most circumstances, if you ever have perfect prediction, or even near perfect prediction, you have either asked a rather obvious question of your data or have accidentally included the target in your features (or a combination of them).

$$y = b_0 + b_1x_1 + b_2x_2 + ... + b_nx_n + \epsilon$$

The error term is a random variable that represents the difference between the actual value and the predicted value.  We can't know what the error term is, but we can estimate it.  We'll talk more about that in the section on Estimation.


### Prediction vs. Explanation

In our view, one can't stress enough the importance of a model's ability to predict the target.  It can be a poor model, maybe because the data is not great, or we're exploring a new area, but we'll always be interested in how well a model **fits** the observed data, and predicts new data.

As strange as it may sound, you can read whole journal articles and business reports in many fields with hardly any mention of prediction. Instead, the focus is on the **explanation** of the model, and this is where the coefficients come in. They are used to explain how the features are related to the target, and we may use terms like saying a feature 'strongly' correlates, or 'negatively' correlates to the target.  In our case, we can say that for every additional word in the review, we *expect* the rating to go up by `r rl_coef` stars.  We might even use our background knowledge and context to say this is a 'strong' relationship.  

Additionally, the extra information in the table shows that review length is **statistically significant**, which is a very loaded description.  For our purposes right now, we'll interpret it as follows: if we expected the coefficient to be zero, the probability of seeing a coefficient as large as `r rl_coef` or larger is very small.  In fact, the probability is so small (rounded to zero), we might conclude that the coefficient is not zero, and that the feature is related to the target.  Unfortunately, statistical significance is affected by other things besides the size of the coefficient, and without an understanding of the context of the features (like how long typical reveiws are, what their range is, what variability of ratings is, etc.), the information it provides is extremely limited, and many would argue, not even useful at all.

In the past, and unfortunately even the present, statistical significance is focused on a great deal, even to the point that a papers are written about models that have no predictive power at all. In those settings, statistical significance is often used as a proxy for importance, which it never should be. If we are very interested in the coefficient, it is better to focus on the range of possible values, which is provided by the **confidence interval**.  While a confidence interval is also a loaded description of a feature's relationship to the target, we can use it in a very practical way as a range of possible values for that weight. Here we see that the coefficient could be as low as `r rl_ci[1]` or as high as `r rl_ci[2]`, which we might interpret as a relatively narrow range, since the , giving us some 'confidence' that we should expect a non-zero value when assessing this particular relationship. Like statistical significance though, it is affected by many things, and without context, is limited in the information it can provide.  

Suffice it to say at this point that how much one focuses on prediction vs. explanation depends on the context and goals of the data endeavor.  There are cases where predictive capability is of utmost importance, and we care less about the explanation, but not to the point of ignoring it. Even with deep learning models for image classification, where the inputs are just RGB values, we'd still like to know what the (notably complex) model is picking up on, otherwise we may be classifying images based on background nonsense.  In some business settings, we are very or even mostly interested in the weights, which might indicate how to allocate resources in some fashion, but if they come from a model with no predictive power, this may be a fruitless endeavor.





### Prediction Error

## How do we obtain a model? Estimation

In our linear model, one of, and so far the only, key **parameter** is the coefficient for each feature. But how do we know what the coefficients are? When we run a linear model using some program function, they appear magically, but it's worth knowing a little bit about how they come to be.


In **model estimation**, we can break things down into the following steps:

1. Start with an initial guess for the parameters
2. Calculate the **prediction error** or some function of it
3. **Update** the guess
4. Repeat steps 2 & 3 until we find a 'best' guess


:::{.callout type="info" title="Estimation vs. Optimization"}
We can use **estimation** as general term for finding parameters, while **optimization** can be seen as a term for finding parameters that maximize or minimize some function.  In some cases we can estimate parameters without optimization, because there is a known way of solving the problem, but in most modeling situations we are going to use some optimization approach.
:::



### Guessing

As a starting point we can just guess what the parameter should be, but how would we know what to guess? How would we know which guesses are better than others?  Let's try a few guesses and see how they do.  We'll start with a guess of 0 for all the coefficients.  We can plug this into the model and see what we get: 

$$\textrm{prediction} = 0 + 0\cdot\textrm{review\_length}$$

But that's just going to equal zero, which we know would not be a good guess.  A better guess would just be the mean of the target.  We can plug that into the model and see what we get:

$$\textrm{prediction} = \bar{rating} + 0\cdot\textrm{review\_length}$$

In this case, our guess for the coefficients are still zero, but our offset (or intercept) is the mean of the target. This is a better guess, and is at least data driven, but it's still not great. How do we know it's better?  

### Prediction Error

We can compare the predictions from each guess to the actual values of the target.  We can do this by calculating the **prediction error**, or in the context of a linear model, they are also called **residuals**.  The prediction error is the difference between the actual value of the target and the predicted value of the target.  We can express this as:

$$ \epsilon = y - \hat{y}$$
$$ \textrm{error} = \textrm{target} - \textrm{(model based) guess}$$

Not only does this tell us how far off our model prediction is, it gives us a way to compare models.  With a measure of prediction error, we can get a **metric** for total, or maybe the average error, and if one model has less total/average error, we can say it's a better model.  Ideally we'd like to choose a model with the least error, but we'll see that this is not always possible[^neverbest].  For now, let's calculate the error for our two guesses. One thing though, if we miss the mark above or below our target, we still want it to count the same in terms of prediction error. In other words, if the true rating is 3 and our model predicts 3.5 or 2.5, we want those to count the same when we total up our error[^absloss]. One way we can do this is to use the squared error, or use absolute value.  We'll use squared error here, and we'll calculate the sum of the squared errors for all our predictions.  We'll do this in R and Python, and then compare the results.

[^absloss]: We don't have to do it this way, but it's the default in most scenarios. As an example, maybe for your situation overshooting is worse than undershooting, and so you might want to use an approach that would weight those errors more heavily.

[^neverbest]: It turns out that our error value is an estimate itself of the true error, and we'll talk more about this later. For now, this means that we can't ever know the true error, and so we can't ever know the best model. We can still get a good or better model relative to others.

:::{panel-tabset}

###### R

```{r}
#| label: r-error
y = df_reviews$rating

# Calculate the error for the guess of all zeros
error_zeros <- sum((y - 0)^2)

# Calculate the error for the guess of the mean
error_mean <- sum((y - mean(y))^2)
```


###### Python

```{python}
#| label: py-error
y = df_reviews['rating']

# Calculate the error for the guess of all zeros
error_zeros = np.sum((y - 0)**2)

# Calculate the error for the guess of the mean
error_mean = np.sum((y - y.mean())**2)

```

:::

```{r}
#| echo: false
#| label: compare-error
tibble(
    zero_sq_sum = error_zeros,
    mean_sq_sum = error_mean
) |>
    gt()
```

Well, this is useful, but you're probably hoping there is an easier way to do this, and there is!


- OBJECTIVE FUNCTION
- LOSS FUNCTION
- COST FUNCTION


### Ordinary (?) Least Squares

For a simple linear model, we can estimate the parameters in several ways, but the most common is to use the **Ordinary Least Squares (OLS)** method. OLS is a method of estimating the coefficients that minimizes the sum of the squared errors, which we've just been doing[^notamodel].  In other words, it finds the coefficients that minimize the sum of the squared differences between the predicted values and the actual values.  We can express this as:

[^notamodel]: Some disciplines seem confuse models with estimation methods and link functions. It doesn't really make sense nor is informative to call something an OLS model or a logit model. Many models are estimated using a least squares approach, and different types of models use a logit link.

$$\textrm{Loss} = \sum_{i=1}^{n} (y_i - \hat{y_i})^2$$

Where $y_i$ is the actual value of the target for observation $i$, and $\hat{y_i}$ is the predicted value from the model.  The sum of the squared errors is also called the **residual sum of squares** (RSS). The OLS method finds the coefficients that minimize the sum of the squared differences between the predicted values and the actual values.

It's called *ordinary* least squares becuase there are other least squares- generalized least squares, weighted least squares, and others, but this doesn't really matter. What matters is that we have a way to estimate the coefficients that minimizes the sum of the squared errors. 

The resulting value - the sum or mean of the squared errors, is sometimes referred to as our **objective** function, or **loss** function (if we are minimizing it), or even target function.  We can use this value to find the best parameters for a specific model, as well as compare different models with different parameters, such as a model with different features.  We can also use this value to compare different types of models, such as a linear model and a decision tree model.  

Let's calculate the OLS estimate for our model. From our steps above, we need guesses and a way to update them.  Well, we can provide a bunch of guesses, and just move along from one set to the next until we find the best one.


:::{.panel-tabset}

##### R
```{r}
#| label: r-ols
ols <- function(x, y, b0, b1, sum = TRUE) {
    # Calculate the predicted values
    y_hat <- b0 + b1 * x
    
    # Calculate the error
    error <- y - y_hat
    
    # Calculate the value as sum or average
    if (sum) {
        value <- sum(error^2, na.rm = TRUE)
    } else {
        value <- mean(error^2, na.rm = TRUE)
    }

    # Return the value
    return(value)
}

# create a grid of guesses
guesses <- crossing(
    b0 = seq(-2, 2, 0.1),
    b1 = seq(-2, 2, 0.1)
)

# Example for one guess
ols(
    x = df_reviews$review_length,
    y = df_reviews$rating,
    b0 = guesses$b0[1],
    b1 = guesses$b1[1]
)

```

##### Python
```{python}
#| label: py-ols
#| 
from itertools import product # to create our grid of guesses

def ols(x, y, b0, b1, sum = True):
    # Calculate the predicted values
    y_hat = b0 + b1 * x
    
    # Calculate the error
    error = y - y_hat
    
    # # Calculate the value as sum or average
    if sum:
        value = np.sum(error**2)
    else:
        value = np.mean(error**2)
    
    # Return the value
    return(value)

# create a grid of guesses

guesses = pd.DataFrame(
    product(
        np.arange(-2, 2, 0.1),
        np.arange(-2, 2, 0.1)
    ),
    columns = ['b0', 'b1']
)


# Example for one guess
ols(
    x = df_reviews['review_length'],
    y = df_reviews['rating'],
    b0 = guesses['b0'][0],
    b1 = guesses['b1'][0],
    sum = False
)
```

:::

Now we want to calculate the loss for each guess and find which one gives us the minimum function value.

```{r}
#| label: r-ols-apply
#| echo: false

# Calculate the ols function value for each guess
guesses <- guesses %>%
    mutate(objective = map2_dbl(
        guesses$b0, guesses$b1,
        \(b0, b1) ols(b0 = b0, b1 = b1, x = df_reviews$review_length, y = df_reviews$rating)
    ))

min_loss = guesses %>% filter(objective == min(objective))

predictions = min_loss$b0 + min_loss$b1 * df_reviews$review_length

guesses |>
    ggplot(aes(x = b0, y = b1)) +
    geom_tile(aes(fill = objective), show.legend = FALSE) +
    geom_point(
        data = min_loss,
        size = 6,
        color = "white"
    ) +
    geom_text(
        data = min_loss,
        aes(label = glue("Minimum at ({round(b0, 2)}, {round(b1, 2)})\nObjective value = {round(objective, 2)}")),
        size = 3,
        hjust = -0.0,
        color = "gray50"
    ) +
    annotate(
        geom = "text",
        x = intercept,
        y = rl_coef,
        label = glue("OLS estimate at ({intercept}, {rl_coef})\nObjective value = {round(summary(model_reviews)$sigma^2*nrow(df_reviews), 2)}"),
        size = 3,
        hjust = -0.1,
        color = "gray50"
    ) +
    # coord_cartesian(
    #     # xlim = c(-1, 1),
    #     ylim = c(-1, 1)
    # ) +
    scico::scale_fill_scico() +
    labs(
        x = "b0",
        y = "b1",
        title = "Objective value (loss) for different guesses of b0 and b1"
    )
```

Returning to our results from the built-in functions, we had estimates of `r intercept` and `r rl_coef` for our coefficients. These are similar but not exactly the same, but we are very close on the weight for review length, and not wildly off on the intercept.

```{r}
p1 = df_reviews |> 
    mutate(prediction = predictions)  |> 
    ggplot(aes(x = prediction, y = rating)) + 
    geom_point(alpha = 0.5, size = 4) +
    geom_abline(
        intercept = 0,
        slope = 1,
        color = okabe_ito[2],
        linewidth = 2
    )

# compare densities
p2 = df_reviews |> 
    mutate(prediction = predictions) |> 
    select(rating, prediction) |>
    pivot_longer(everything()) |>
    ggplot(aes(x = value)) +
    geom_density(aes(fill = name), alpha = 0.5) 

p1  + p2
```




### Estimation as 'Learning'




### Maximum Likelihood 



### Gradient Descent?

### Other Loss Functions Quick Demo


## Assumptions and More

## Commentary

- Opinions
- Limitations/Failure points
- Summary