# Miscellaneous Terminology

We saw several ways to refer to features/inputs and targets/outputs in @tbl-feature-target-names. Here are some more to consider.

```{r}
#| echo: false
#| label: tbl-sample-names
#| tbl-cap: Common Terms for Observations

tbl_observations = tibble(
    Name = c("data", "observation", "sample", "instance", "example"),
) |>
    gt()
```

When we want to use data to assess performance, we will use splits of the data. Here are some common terms for these splits. For each, we'll note their primary use initial model assessment (training) or not (testing). To be clear, sets used in the (cross-) validation process are still used in training whenever they aren't being used for validation.

```{r}
#| echo: false
#| label: tbl-split-names
#| tbl-cap: Common Terms for Data Splits

tbl_splits = tibble(
    Name = c("train", "validation", "test", "holdout", "out-of-sample"),
    Type = c('training', 'training', 'testing', 'testing', 'either')
) |>
    gt()
```

Some may refer to out-of-sample data as holdout data. This is because it is held out of the original training sample of data and/or was never included. However, it can also be used for validation, for example, within a random forest model, typically called the out-of-bag data.


