---
title: "Untitled"
format: html
---



```{r}
library(tidyverse)
reticulate::use_condaenv("book-of-models")
```


## Movie Reviews

TODO: 
- Get positive correlations back
- More variability in review length? Still cap for ease.
- Limit range of review year


Movie based:
- title  (at least 50 unique titles)
- year of release (positive correlation with rating; minimum year 1980, maximum year 2020)
- length minutes  (mostly positive correlation with rating, minimum 60, maximum 180)
- genre (higher ratings in for drama, lower ratings for comedy no more than 6 genres)
- season released (higher ratings in summer)
- total number of reviews given to the movie (slight positive correlation with rating, minimum 100, maximum 1000)

Reviewer based:
- age (negative correlation with rating)
- number of children at home (higher correlation for kids movies)
- work status (no relationship with rating)
- education level (3 levels- completed high school, completed college, post-graduate; slightly negative relationship with rating)
- gender (no relationship with rating)

Review based:
- the actual review text (word count minimum 2, maximum 25; word count should generally be positively correlated with rating)
- review year (higher ratings in more recent years)


- review length (positive correlation with rating)
 - if word count we'd probably want a bit more variability
- review complexity
- sentiment

** GPT Prompt for Titles **

Please provide titles appropriate to the Drama genre. Present results in a table with the following columns:
title
genre


30 titles from the Drama genre
27 titles from Action/Adventure
10 titles from the Comedy genre
10 titles from Romance
8 titles for Kids movies
5 SciFi titles
5 Horror Titles
5 misc titles (e.g. documentary)

** GPT Prompt for Text **

Initial:
Provide 100 random positive/neutral/negative movie reviews.

Length based:
- for positive, we add 30 short (3-5 words), and 20 long reviews
- for negative, we add 10 short (3-5 words), and 40 long reviews

Note that trying to do both short and long confused ChatGPT.
'Provide x random positive/negative movie reviews. They should be at least fifteen words in length.'

'Provide 50 random neutral movie reviews. # neutral only added for balance'



## Data Generation Setup

```{r} 
#| label: import-ready-made-data
positive_reviews = unique(readr::read_tsv("data/positive.tsv")[[1]])
neutral_reviews = unique(readr::read_tsv("data/neutral.tsv")[[1]])
negative_reviews = unique(readr::read_tsv("data/negative.tsv")[[1]])
movie_titles = read_tsv("data/movie_titles.tsv") # two columns title and genre
```


```{r}
#| label: functions
generate_review = function(rating) {
    reviews = case_when(
        rating >= 3.5  ~ sample(positive_reviews, replace = TRUE, length(rating)),
        rating <= 2.75 ~ sample(negative_reviews, replace = TRUE, length(rating)),
        TRUE ~ sample(neutral_reviews, replace = TRUE, length(rating))
    )

    return(reviews)
}
    

generate_data = function(n, seed = 42) {
    set.seed(seed) 

    ## Reviewer Data
    df_reviewer = tibble(
        title = sample(movie_titles$title, n, replace = TRUE),
        review_year = sample(2000:2022, n, replace = TRUE, prob = (2000:2022 - 1999)^1.5),
        age = rnorm(n, 45, 20),
        # kids will basically be capped at 2+ 
        children_in_home = case_when(
            age >= 55 & age < 65 ~ sample(
                0:3, 
                n, 
                replace = TRUE, 
                prob = c(0.90, 0.09, .01, 0)
            ),
            age >= 65 ~ sample(
                0:3, 
                n, 
                replace = TRUE, 
                prob = c(0.99, 0.01, 0, 0)
            ),
            age <= 25 ~ sample(
                0:3, 
                n, 
                replace = TRUE, 
                prob = c(0.90, 0.09, 0.01, 0)
            ),
            age > 25 & age < 55 ~ sample(
                0:3, 
                n, 
                replace = TRUE, 
                prob = c(0.4, 0.45, 0.1, 0.05)
            ),
            TRUE ~ 0
        ),
        education = sample(c("Completed High School", "Completed College", "Post-Graduate"), n, replace = TRUE, prob = c(.4, .6, 1)),
        gender = sample(c('male', 'female'), n, replace = TRUE, prob = c(0.4, 0.6)),
        # hidden variables 
        reviewer_type = case_when(
            age <= 25 ~ sample(c('Critic', 'Viewer'), n, replace = TRUE, prob = c(0.1, 0.9)),
            age >= 35 ~ sample(c('Critic', 'Viewer'), n, replace = TRUE, prob = c(0.3, 0.7)),
            TRUE ~ sample(c('Critic', 'Viewer'), n, replace = TRUE, prob = c(0.2, 0.8))
        ),
        # random_x = rnorm(n, 0, 1),
    ) |> 
    mutate(
        age = ifelse(age < 18, 18, round(age)),
        age = ifelse(age > 80, 80, age),
        work_status = case_when(
            age <= 25 ~ sample(
                c('Employed', 'Unemployed', 'Student', 'Retired'), 
                n, 
                replace = TRUE, 
                prob = c(0.5, 0.1, 0.4, 0.0)
            ),
            age >= 70 ~ sample(
                c('Employed', 'Unemployed', 'Student', 'Retired'), 
                n, 
                replace = TRUE, 
                prob = c(0.2, 0.1, 0.0, 0.7)
            ),
            TRUE ~ sample(
                c("Employed", "Unemployed", "Student", "Retired"), 
                n, 
                replace = TRUE, 
                prob = c(0.8, 0.05, 0.02, 0.04)
            )
        ),
    ) 

    ## Movie Data
    n_movies = n_distinct(df_reviewer$title)

    df_movies = movie_titles |> inner_join(df_reviewer |> distinct(title), by = 'title')

    df_movies = df_movies |> 
        mutate(
            release_year = sample(1980:2020, n_movies, replace = TRUE, prob = (1980:2020 - 1979)^1.5),
            length_minutes = rpois(n_movies, 120), 
            season = sample(c('Winter', 'Spring', 'Summer', 'Fall'), n_movies, replace = TRUE, prob = c(1, 1, 3, 2)), 
            total_reviews = sample(100:10000, n_movies, replace = TRUE)
        )

    df = df_reviewer |>
        left_join(df_movies, by = 'title') 

    # ratings and review text features
    # main effects here will allow for whole changes (e.g. doubling) or 
    # otherwise just add organization
    main_effects = tibble(
        release_year   = .2, 
        review_year    = .1, 
        length_minutes = .3,
        length_minutes_sq = -.1, 
        drama = .5, 
        comedy = -.3,
        summer = .2,
        total_reviews = .5,
        age = -.1,
        children_in_home = .25,
        student = .25,
        post_graduate = -.1
    )

    multiplier = 2
    interact_mult = multiplier * .5
    main_effects = main_effects*multiplier

    rating = (
        main_effects$release_year * scale(df$release_year)
        + main_effects$review_year * scale(df$review_year)
        + main_effects$length_minutes * scale(df$length_minutes) - main_effects$length_minutes_sq * scale(df$length_minutes)^2 # will have to tinker with this
        + main_effects$drama * scale(df$genre == 'Drama')
        - main_effects$comedy * scale(df$genre == 'Comedy')
        + main_effects$summer * scale(df$season == 'Summer')
        + main_effects$total_reviews * scale(df$total_reviews)
        - main_effects$age * scale(df$age) 
        + main_effects$children_in_home * scale(df$children_in_home > 0)
        + main_effects$student * scale(df$work_status == 'Student')
        - main_effects$post_graduate * scale(df$education == 'Post-Graduate')
        # simple interactions and hidden effects
        # split the data to more high vs. low (but not too much )
        + .5 * scale(df$reviewer_type == 'Viewer')
        - .5 * scale(df$reviewer_type == 'Critic')
        # genre interactions
        + .25 * interact_mult * scale(df$genre == 'Kids' & (df$children_in_home > 0))
        + .1 * interact_mult * scale(df$genre %in% c('Romance', 'Horror') & (df$age < 30))
        + .1 * interact_mult * scale(df$genre == 'Comedy' & df$work_status == 'Student')
        + rnorm(nrow(df), 0, 1) 
    ) 
    
    df = df |> 
        mutate(
            rating = round(scales::rescale(rating[, 1], to = c(1, 5)), 1),
            review_text = generate_review(rating),
            word_count  = str_count(review_text, '\\w+')
        ) |> 
        select(-reviewer_type)
    df
}
```

## Generate, Explore and Test

```{r}
df = generate_data(1000, seed = 42)
# df$review_text
# glimpse(df)
# df |> count(review_text, sort = TRUE)
# hist(df$rating)
# cor(df$rating, df$word_count)
corrr::correlate(df) |>
    autoplot()
```

Misc checks
```{r}
skimr::skim(df)
df |>
    mutate(age = cut(age, breaks = c(0, 25, 55, 65, 100))) |>
    count(age, children_in_home)
skimr::skim(df |> group_by(genre) |> select(rating)) |> gt()
```


```{r}
test_mod = lm(
    rating ~
        release_year
        + review_year
        + length_minutes
        + genre
        + season
        + total_reviews
        + age
        + children_in_home
        + work_status
        + education
        + word_count
        + genre:age
        + genre:children_in_home
        + genre:work_status,
    data = df
)

broom::tidy(test_mod) |>
    arrange(desc(abs(statistic))) |>
    print(n = Inf)

summary(test_mod)$r.sq
```


Explore nonlinear relationships here.

```{r}
# library(mgcv)
# d = gamSim(1, n = 100, dist = "normal", scale = 1)
# m = gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = d)
# plot(m, pages = 1)
```



```{r}
pd = tibble(
    x = rpois(1000, 100),
    x_sc = scale(x)[, 1],
    y1 = .2 * x_sc - .05 * x_sc^2,
    y2 = .1 * x_sc - .025 * x_sc^2,
    y3 = .2 * x_sc - .075 * x_sc^2,
    y4 = .3 * x_sc - .05 * x_sc^2
) |>
    select(-x_sc) |>
    pivot_longer(-x)
pd |>
    ggplot(aes(x, value, color = name)) +
    geom_line() +
    geom_smooth(method = "gam") +
    geom_point(data = pd |> summarize(x = x[max(value)], value = max(value), .by = name), color = "red")
```


```{r}
ggplot(df, aes(length_minutes, rating)) +
    geom_point() +
    geom_smooth(method = "gam")
```

## Feature Processing

```{r}
# Subtract minimum from review year, release year
# Scale age, length minutes, total reviews, word count, character count
# create a binary rating for >=3 equals good, <3 equals bad

df_processed <- df |>
    mutate(
        review_year_0 = review_year - min(review_year),
        release_year_0 = release_year - min(release_year),
        age_sc = scale(age)[, 1],
        length_minutes_sc = scale(length_minutes)[, 1],
        total_reviews_sc = scale(total_reviews)[, 1],
        word_count_sc = scale(word_count)[, 1],
        # character_count_sc = scale(character_count)[, 1],
        rating_good = ifelse(rating >= 3, 1, 0)
    )

```

## Save
```{r}
write_csv(
    df, 
    "data/movie_reviews.csv"
)
write_csv(
    df_processed, 
    "data/movie_reviews.csv" # no need to save as separate objects
)
```